---
title: "README"
author: "Lukas DiGiovanni"
date: "2025-04-25"
output:
 github_document:
 pandoc_args: ["--wrap=none"]
always_allow_html: true
---

```{r setup, include=FALSE}
# Set working directory
setwd("/Users/lukasdigiovanni/Desktop/EDA_withR")
```

# Exploratory Data Analysis with R- Capstone Deliverable

### By Lukas DiGiovanni

# Wine Quality Analysis

## Introduction

Wine quality is influenced by a variety of chemical and physical
properties, many of which can be measured objectively through laboratory
analysis. Factors such as alcohol content, acidity levels (fixed,
volatile, and citric), residual sugar, pH, sulphates, and sulfur dioxide
concentrations are all known to significantly impact the taste, aroma,
and overall acceptability of wine.

Higher quality wines tend to exhibit: - Higher alcohol content
(associated with better body and aroma), - Balanced acidity (not too
sour or flat), - Lower volatile acidity (excess can produce an
unpleasant vinegar taste), - Higher levels of sulphates (which help
preserve the wine and enhance flavor stability), - Appropriate sugar
levels depending on the wine style.

Conversely, elevated levels of residual sugar without proper acidity can
make wine seem cloying, while high densities and certain forms of
acidity can negatively affect mouthfeel.

This analysis combines two datasets — one for red wine and one for white
wine — and introduces a binary variable (`red`) to distinguish between
the two types. Throughout this project, we will explore how these
different chemical properties relate to perceived wine quality and
whether patterns differ between red and white wines.

Future visualizations may include: - Comparing distributions of alcohol
and acidity across wine types, - Exploring relationships between
chemical features and quality scores, - Building models to predict wine
quality based on physicochemical attributes.

By understanding which features are most predictive of quality, we can
gain insights into the art and science behind winemaking excellence.

------------------------------------------------------------------------

```{r load-data}
# Load the two CSV files
# Load the two datasets
red_wine <- read.csv("winequality-red.csv", sep = ";")
white_wine <- read.csv("winequality-white.csv", sep = ";")

# Add a new column to each indicating color
red_wine$color <- "red"
white_wine$color <- "white"

# Combine the datasets
wine_data <- rbind(red_wine, white_wine)
```

## Data Understanding

Below is a summary of the combined dataset `wine_data`, including the
number of samples, number of features, and a breakdown of each feature's
type and description.

```{r}
# Number of rows (samples) and columns (features)
num_samples <- nrow(wine_data)
num_features <- ncol(wine_data)

cat("Number of samples:", num_samples, "\n")
cat("Number of features:", num_features, "\n")

```

```{r}
# Structure of the dataset: data types
str(wine_data)

```

```{r}
# Create a data frame with feature metadata
feature_info <- data.frame(
  Feature = names(wine_data),
  Description = c(
    "Fixed acidity (tartaric acid)",
    "Volatile acidity (acetic acid)",
    "Citric acid (flavor enhancer)",
    "Residual sugar (g/L)",
    "Chlorides (salt content)",
    "Free sulfur dioxide (preservative)",
    "Total sulfur dioxide (preservative)",
    "Density of the wine",
    "pH (acidity/alkalinity)",
    "Sulphates (preservative, flavor)",
    "Alcohol percentage",
    "Quality rating (0–10 scale)",
    "Wine type (1 = red, 0 = white)"
  ),
  Type = c(
    "Ratio", "Ratio", "Ratio", "Ratio", "Ratio", "Ratio",
    "Ratio", "Ratio", "Interval", "Ratio", "Ratio", "Ordinal", "Nominal"
  )
)

# Display the table nicely
library(knitr)
kable(feature_info, caption = "Feature Descriptions and Data Types")

```

```{r}
## Summary Statistics

# Basic summary of the dataset
summary(wine_data)

# If you want a cleaner summary for only numeric variables:
numeric_cols <- sapply(wine_data, is.numeric)
summary(wine_data[, numeric_cols])

# Number of observations and features
cat("Number of observations:", nrow(wine_data), "\n")
cat("Number of features:", ncol(wine_data), "\n")
```

# EDA

# ------------------------------------------------------------------------------
## Correlation of Numerical Features with Wine Quality
# ------------------------------------------------------------------------------
```{r}
# Look at correlation with quality
correlations <- cor(wine_data[sapply(wine_data, is.numeric)])
round(correlations["quality", ], 2)

```
### Interpretation:
 - The variable most strongly **positively correlated** with wine quality is **alcohol (r = 0.44)**.
   This supports the earlier hypothesis that higher alcohol content is a key quality driver.
 - Other weakly positive correlations include **citric acid (r = 0.09)** and **free sulfur dioxide (r = 0.06)**,
   though these are very small and may not be practically significant.
 - Several variables are **negatively correlated** with quality:
   - **Density (r = -0.31)**: More dense wines tend to have lower quality.
   - **Volatile acidity (r = -0.27)**: Higher volatility (often associated with sharp or unpleasant flavors) reduces quality.
   - **Chlorides (r = -0.20)**: High salt content seems to lower quality.
   - **Fixed acidity (r = -0.08)** and **total sulfur dioxide (r = -0.04)** also show slight negative correlations.
 - **pH (r = 0.02)** and **sulphates (r = 0.04)** are effectively uncorrelated with quality in this dataset.


# ------------------------------------------------------------------------------
## Alcohol Content by Wine Quality (Boxplot)
# ------------------------------------------------------------------------------


```{r}
library(ggplot2)
# Convert quality to factor for proper boxplot grouping
wine_data$quality <- as.factor(wine_data$quality)

# Recreate alcohol vs. quality boxplot
ggplot(wine_data, aes(x = quality, y = alcohol)) +
  geom_boxplot() +
  labs(title = "Alcohol Content by Wine Quality", y = "Alcohol", x = "Quality")
```

# Question: Does alcohol content affect wine quality?
```{r}
library(ggplot2)

# Alcohol vs Quality
ggplot(wine_data, aes(x = quality, y = alcohol)) +
  geom_boxplot() +
  labs(title = "Alcohol Content by Wine Quality", y = "Alcohol", x = "Quality")

```
 Hypothesis: Wines with higher alcohol content tend to have higher quality.

### Interpretation:
 - This boxplot shows how alcohol levels vary across wine quality ratings from 3 to 8.
 - The **median alcohol content** clearly increases with wine quality.
 - Wines rated 7 and 8 have **higher median and upper quartile alcohol content** than wines rated 3 to 5.
 - Lower quality wines (quality 3–5) tend to have a narrower and lower alcohol distribution.
 - A few outliers are observed (especially in quality 6), but the trend overall is upward.
 - This visualization provides **strong visual support** for the earlier correlation result (r = 0.44)
   indicating a **moderately strong positive relationship** between alcohol and quality.

### Why this matters:
 Alcohol is a significant chemical component that can affect mouthfeel, flavor, and perceived richness, attributes likely to influence quality ratings.

# ------------------------------------------------------------------------------
## Boxplot: Volatile Acidity by Wine Quality
# ------------------------------------------------------------------------------

```{r}

# Volatile Acidity vs Quality
ggplot(wine_data, aes(x = quality, y = volatile.acidity)) +
  geom_boxplot() +
  labs(title = "Volatile Acidity by Wine Quality", y = "Volatile Acidity", x = "Quality")

```
### Hypothesis:
 Wines with higher volatile acidity are rated lower in quality.

### Interpretation:
 - The plot shows a clear **negative relationship** between volatile acidity and wine quality.
 - As wine quality increases, the **median volatile acidity decreases**.
 - Higher quality wines (7–8) show **lower typical volatile acidity** than lower quality wines (3–5).
 - A large number of outliers exist—especially in quality level 6—indicating variability, but the trend remains.
 - This supports the correlation result from earlier (r = -0.27), indicating a **moderate negative correlation**.

### Why this matters:
 Volatile acidity reflects the concentration of acetic acid (vinegar-like taste).
 Excessive volatile acidity is often perceived as a fault in wine, which explains why wines with lower levels are typically rated higher in quality.

# ------------------------------------------------------------------------------
## Sulphates by Wine Quality (Boxplot)
# ------------------------------------------------------------------------------

### Question: Are sulphate levels associated with higher wine quality?

```{r}
# Sulphates vs Quality
ggplot(wine_data, aes(x = quality, y = sulphates)) +
  geom_boxplot() +
  labs(title = "Sulphates by Wine Quality", y = "Sulphates", x = "Quality")
```

### Hypothesis:
 Higher sulphate levels are associated with higher wine quality.

### Interpretation:
 - The median sulphate level shows a **slight increasing trend** with wine quality.
 - However, the differences between boxplots are **subtle** compared to alcohol or volatile acidity.
 - Most wines have sulphate levels between 0.4 and 0.7, regardless of quality.
 - A large number of **outliers are present**, particularly in the mid-range (quality 6), indicating variability.
 - This supports the earlier correlation result of **r = 0.04**, which is **very weak**.
   It suggests **no meaningful linear relationship** between sulphates and quality.

### Why this matters:
 While sulphates (used as preservatives) can affect flavor stability and microbial resistance,
 their influence on perceived quality may be marginal in comparison to other chemical properties like alcohol..


# ------------------------------------------------------------------------------
## T-Test: Alcohol vs. Quality Group
# ------------------------------------------------------------------------------

```{r}
# Ensure 'quality' is numeric before comparison
wine_data$quality <- as.numeric(as.character(wine_data$quality))

# Grouping wine into high vs low quality
wine_data$quality_group <- ifelse(wine_data$quality >= 7, "high", "low")

# Confirm it worked (optional but helpful)
table(wine_data$quality_group)

# Run Welch's t-test for alcohol across the two quality groups
t.test(alcohol ~ quality_group, data = wine_data)

```

## Assumptions

### Normality

```{r}
set.seed(42)
sample_high <- sample(wine_data$alcohol[wine_data$quality_group == "high"], size = 500)
sample_low  <- sample(wine_data$alcohol[wine_data$quality_group == "low"], size = 500)

shapiro.test(sample_high)
shapiro.test(sample_low)
```

#### Results
 - Shapiro-Wilk tests for normality yielded significant p-values for both high- and low-quality wine groups, indicating departures from perfect normality. However, due to the large sample sizes and the t-test's robustness to normality violations, especially under Welch’s correction, the test results are considered valid.

### Equality of Variances

```{r}
library(car)
leveneTest(alcohol ~ quality_group, data = wine_data)
```

#### Results
 - Levene’s test for homogeneity of variance yielded a highly significant result (F = 36.26, p < 0.001), indicating unequal variances between high- and low-quality wine groups. This supports the use of Welch’s t-test, which does not assume equal variances.

### Interpretation
 - Normality tests indicate deviations in both groups, but Welch’s t-test is robust under large sample sizes. Levene’s test confirms unequal variances, further validating the choice of Welch’s t-test. Independence is assumed based on data collection. Therefore, the assumptions necessary for valid inference are sufficiently met.
 
## Hypothesis: Mean alcohol content is higher in high-quality wines.
### Assumptions:
 - Independent groups (high vs. low)
 - Roughly normal distribution of alcohol in each group
 - Variances may differ; Welch's t-test handles this
 Interpretation: If p-value < 0.05, alcohol is significantly different between high and low quality groups, supporting our hypothesis.
 
### Result
 - There is a statistically significant difference in alcohol content between high- and low-quality wines (p < 2.2e-16). On average, high-quality wines have an alcohol content of 11.43%, compared to 10.26% in low-quality wines — a mean difference of ~1.17%, with a 95% confidence interval of [1.10%, 1.24%]. These findings support the hypothesis that higher alcohol content is associated with better wine quality.

# ------------------------------------------------------------------------------
## Alcohol Distribution by Quality Group (Histogram)
# ------------------------------------------------------------------------------


```{r}
# Check normality with histograms
ggplot(wine_data, aes(x = alcohol)) +
  geom_histogram(bins = 30, fill = "skyblue") +
  facet_wrap(~ quality_group) +
  labs(title = "Alcohol Distribution by Quality Group")

# Check variance
library(car)
leveneTest(alcohol ~ quality_group, data = wine_data)

```
### Hypothesis: Higher quality wines are more likely to have higher alcohol.
Interpretation: High-quality wines cluster around 11–13% alcohol.
Low-quality wines peak at lower alcohol values (~9–10%).
Visually supports the hypothesis that alcohol differentiates quality.

# ------------------------------------------------------------------------------
## Correlation Heatmap
# ------------------------------------------------------------------------------


```{r}
# Correlation matrix heatmap
library(corrplot)
corrplot(cor(wine_data[sapply(wine_data, is.numeric)]), method = "color", type = "upper")

# Simple linear model: Alcohol predicting Quality
model <- lm(quality ~ alcohol, data = wine_data)
summary(model)

```
## Interpretation:
Alcohol shows the strongest positive correlation with quality (~0.45).
Sulphates and citric acid show weaker positive correlations.
Density is negatively correlated with quality.
Alcohol is the most promising predictor for wine quality.


# Simple linear model: Alcohol predicting Quality
```{r}
wine_data$quality <- as.numeric(as.character(wine_data$quality))  # safely convert factor to numeric
model <- lm(quality ~ alcohol, data = wine_data)
summary(model)

```

## Checkiing Assumptions

### Linearity

#### Residuals vs Fitted
```{r}
plot(model, which = 1)  # Residuals vs Fitted
```

#### Results
 - There is slight curviture, but it maintains a mostly flat loess line.
 - The residuals appear evenly distributed around 0.
 - Minor structure due to discrete quality scores, but does not violate linearity.


## Normality

### Q-Q Plot

```{r}
# Q-Q plot to visually assess normality
plot(model, which = 2)
```

#### Results
 - The Q-Q plot shows that residuals are approximately normal in the central range, with moderate deviations in the tails. Given the large sample size, these deviations are not severe enough to violate the normality assumption for inference purposes.

##Independence of Residuals

```{r}
library(lmtest)
dwtest(model)  # Durbin-Watson test

```

#### Results
 - The Durbin-Watson test yielded a statistic of 1.63 with a p-value < 2.2e-16, indicating significant positive autocorrelation in the residuals. This suggests that the assumption of independent errors is violated, which may be due to omitted variables or structure in the data. While the core trend between alcohol and quality remains informative, this limitation should be noted when interpreting the model's inference results.

```{r}
plot(model, which = 4)  # Cook’s distance

```

#### Results
 - Cook’s Distance was calculated to assess influential observations. While a few data points (e.g., obs 653, 2340, and 5501) exhibited relatively higher influence, all values remained well below the common threshold of 0.5, indicating no strong outliers affecting the model.
 

# Multiple Linear Regression
```{r}
# Fit improved multiple regression model with polynomial, interaction, and extra predictors
model_improved <- lm(
  quality ~ poly(alcohol, 2) * sulphates +
    volatile.acidity +
    citric.acid +
    color +
    fixed.acidity +
    residual.sugar +
    pH +
    density +
    free.sulfur.dioxide +
    total.sulfur.dioxide +
    chlorides,
  data = wine_data
)

# View summary
summary(model_improved)

```

## Assumptions

### Linearity
```{r}
# 1. Residual plot for linearity + homoscedasticity
plot(model_improved, which = 1)
```

#### Results
 - This returns a flatter loess line, indicating improved linearity.
 
### Normality
```{r}
plot(model_improved, which = 2)
```

#### Results
 - The model still meets the normality assumption, although there are slight skews on the edges.
 
### Autocorrelation
```{r}
library(lmtest)
dwtest(model_improved)
```

#### Results
 - The improved regression model explained more variance in wine quality and included polynomial and interaction terms. However, the Durbin-Watson test (DW = 1.65, p < 0.001) still indicated residual autocorrelation. This suggests that there may be latent structure in the data not captured by the available predictors, or that linear regression may not fully reflect the ordinal nature of wine quality scores.

## Conclusions

Based on our EDA, we observed that alcohol and sulphates are positively correlated with wine quality, while volatile acidity has a negative correlation. Statistical testing (t-tests and linear regression) confirms that higher alcohol content is significantly associated with higher-rated wines. These findings align with real-world winemaking intuition: stronger wines often feel more full-bodied and complex, affecting perception of quality. While there are some issues with the assumptions of the statistical tests, running several tests of varying complexities proved difficult. Both models yield similar DW statistics because the autocorrelation issue is likely due to unobserved grouping or time effects that aren’t captured by your predictors. Since those structural factors aren't included in the dataset, the pattern in residuals remains the same — even in more complex models.

